---
title: "Creating custom annotations"
output: html_document
---

```{r opts}
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618,
  fig.retina = 2, dpi = 150
)
```

```{r packages}
library(tidyverse)
library(here)
library(WDI) # Get data from the World Bank
library(ggrepel) # For non-overlapping labels
library(ggtext) # For fancier text handling
library(showtext)

#bringing the fun fonts in
font_add_google("Schoolbell", "bell") 
font_add_google("Miltonian", "miltonian")

showtext_auto() #load fonts

```

## Annotations

For this example, we're going to use cross-national data from the [World Bank's Open Data portal](https://data.worldbank.org/).

### Load data

```{r load-data-wdi-real}
wdi_co2_raw <- read_csv(here("data", "wdi_co2.csv"))
```

Then we clean the data by removing non-country countries and renaming some of the columns.

```{r clean-data-wdi}
wdi_clean <- wdi_co2_raw %>%
  filter(region != "Aggregates") %>%
  select(iso2c, iso3c, country, year,
    population = SP.POP.TOTL,
    co2_emissions = EN.ATM.CO2E.PC,
    gdp_per_cap = NY.GDP.PCAP.KD,
    region, income
  )
```

### Clean and reshape data

Next we'll do some substantial filtering and reshaping so that we can end up with the rankings of CO~2~ emissions in 1995 and 2014. I annotate as much as possible below so you can see what's happening in each step.

```{r rearrange-data}
co2_rankings <- wdi_clean %>%
  # Get rid of smaller countries
  filter(population > 200000) %>%
  # Only look at two years
  filter(year %in% c(1995, 2014)) %>%
  # Get rid of all the rows that have missing values in co2_emissions
  drop_na(co2_emissions) %>%
  # Look at each year individually and rank countries based on their emissions that year
  group_by(year) %>%
  mutate(ranking = rank(co2_emissions)) %>%
  ungroup() %>%
  # Only select a handful of columns, mostly just the newly created "ranking"
  # column and some country identifiers
  select(iso3c, country, year, region, income, ranking) %>%
  # Right now the data is tidy and long, but we want to widen it and create
  # separate columns for emissions in 1995 and in 2014. pivot_wider() will make
  # new columns based on the existing "year" column (that's what `names_from`
  # does), and it will add "rank_" as the prefix, so that the new columns will
  # be "rank_1995" and "rank_2014". The values that go in those new columns will
  # come from the existing "ranking" column
  pivot_wider(names_from = year, names_prefix = "rank_", values_from = ranking) %>%
  # Find the difference in ranking between 2014 and 1995
  mutate(rank_diff = rank_2014 - rank_1995) %>%
  # Remove all rows where there's a missing value in the rank_diff column
  drop_na(rank_diff) %>%
  # Make an indicator variable that is true of the absolute value of the
  # difference in rankings is greater than 25. 25 is arbitrary here—that just
  # felt like a big change in rankings
  mutate(big_change = ifelse(abs(rank_diff) >= 25, TRUE, FALSE)) %>%
  # Make another indicator variable that indicates if the rank improved by a
  # lot, worsened by a lot, or didn't change much. We use the case_when()
  # function, which is like a fancy version of ifelse() that takes multiple
  # conditions. This is how it generally works:
  #
  # case_when(
  #  some_test ~ value_if_true,
  #  some_other_test ~ value_if_true,
  #  TRUE ~ value_otherwise
  # )
  mutate(better_big_change = case_when(
    rank_diff <= -25 ~ "Rank improved",
    rank_diff >= 25 ~ "Rank worsened",
    TRUE ~ "Rank changed a little"
  ))
```

Here's what that reshaped data looked like before:

```{r show-head-original}
head(wdi_clean)
```

And here's what it looks like now:

```{r show-head-new}
head(co2_rankings)
names(co2_rankings) #variable names
```

### Plot the data and annotate
For we will work together to iteratively build a plot. 

```{r build-pretty-plot, fig.width=10, fig.height=6}
set.seed(123)

ggplot(
  data = co2_rankings,
  mapping = aes(x = rank_1995, y = rank_2014)
) +
  annotate(geom = "segment", x = 0, xend = 172, y = 0, yend = 178) +
  geom_point(aes(color = better_big_change)) +
  geom_label_repel(
    data = filter(co2_rankings, big_change == TRUE),
    aes(label = country, fill = better_big_change),
    color = "white", family = "miltonian"
  ) +
  annotate(
    geom = "text", x = 170, y = 6, label = "Outliers improving",
    family = "bell", fontface = "italic", hjust = 1, color = "grey50"
  ) +
  annotate(
    geom = "text", x = 2, y = 170, label = "Outliers worsening",
    family = "bell", fontface = "italic", hjust = 0, color = "grey50"
  ) +
  scale_color_manual(values = c("grey50", "#0074D9", "#FF4136")) +
  scale_fill_manual(values = c("#0074D9", "#FF4136")) +
  scale_x_continuous(expand = c(0, 0), breaks = seq(0, 175, 25)) +
  scale_y_continuous(expand = c(0, 0), breaks = seq(0, 175, 25)) +
  labs(
    x = "Rank in 1995", y = "Rank in 2014",
    title = "Changes in CO<sub>2</sub> emission rankings between 1995 and 2014",
    subtitle = "Countries that <span style='color: #0074D9'>**improved**</span> or <span style='color: #FF4136'>**worsened**</span> more than 25 positions in the rankings highlighted",
    caption = "Source: The World Bank.\nCountries with populations of less than 200,000 excluded."
  ) +
  guides(color = "none", fill = "none") +
  theme_bw(base_family = "bell") +
  theme( 
    plot.title = element_markdown(face = "bold", size = rel(1.5)), 
    plot.subtitle = element_markdown(size = rel(1.1)), 
    plot.margin = unit(c(0.5, 1, 0.5, 0.5), units = "lines") 
  ) 

```
